---
title: 'Forecasting Canyon Lake Storage: Why ARIMA Outperforms for Water Supply Planning'
author: "Amanda Marrero"
date: "March 22, 2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_selections: true
    df_print: paged
    code_folding: show
    fig_width: 9
    fig_height: 5
    fig_caption: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  dpi       = 150)
```

# Introduction

Canyon Lake, located in Comal County, Texas, serves as a primary surface water 
source for communities across the greater Central Texas region. The Guadalupe-
Blanco River Authority (GBRA) administers the water for wholesale, and the 
reservoir's conservation levels directly govern water availability for municipal 
supply contracts throughout the Guadalupe River Basin. The city of Kyle 
currently receives 54% of its total permitted water supply from Canyon Reservoir, 
but contractually wholesale customers are subject to curtailments across six 
drought response stages, each triggered by declining reservoir levels. Although 
the reservoir has a capacity of 123 billion gallons, persistent drought 
conditions have significantly reduced current storage levels, increasing the 
risk of supply restrictions(City of Kyle, 2025). 
The city of Kyle has experienced significant growth over the last 25 years, 
increasing from 5,000 residents in 2000 to 50,000 in 2020, with projections 
reaching 150,000 by 2080. Kyle remained under Stage 3 drought restrictions 
through 2025, and as the city pursues continued economic expansion, accurate 
long-term water supply forecasting has become an operational necessity (KXAN 
Austin, 2025).
Time series forecasting is a well-established quantitative tool for resource 
planning, with applications ranging from energy demand to reservoir storage 
management. Das et al. (2023) demonstrated that time series models applied to 
Texas reservoirs provide water planners improved ability to make informed 
decisions about water usage, minimizing impacts to local communities and 
businesses. This project applies four forecasting methodologies — Exponential 
Smoothing (ETS), ARIMA, Time Series Linear Regression (TSLM), and an Ensemble 
average — to monthly Canyon Lake storage data spanning January 2010 through 
December 2024, following the modeling framework described in Hyndman & 
Athanasopoulos (2021). The goal is to identify the model best suited for 
supporting water supply planning decisions in a rapidly urbanizing region under 
recurring drought pressure.


# Setup & Libraries

```{r libraries}
library(tidyverse)
library(fpp3)
library(patchwork)
```

```{r color palette}
# Canyon Lake color palette — water/canyon theme
model_colors <- c(
  "ETS_Auto"   = "#2E86AB",   # steel blue
  "ARIMA_Auto" = "#A23B3B",   # clay red  ← selected model gets warmest color
  "TSLM"       = "#3D7A5C",   # canyon green
  "Ensemble"   = "#7B6FA0"    # muted purple
)

model_labels <- c(
  "ETS_Auto"   = "ETS (Automatic)",
  "ARIMA_Auto" = "ARIMA (Automatic)",
  "TSLM"       = "TSLM (Linear Regression)"
)

# Theme
theme_canyon <- function(accent_color = "#2E86AB") {
  theme_minimal() +
    theme(
      plot.title       = element_text(size = 11, face = "bold",
                                      color = "#1a3a2a"),
      plot.subtitle    = element_text(size = 9,  color = "#555555"),
      plot.caption     = element_text(size = 8,  color = "gray50", hjust = 0),
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F7FAFA", color = NA),
      panel.grid.major = element_line(color = "#DDE8E8", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      axis.title       = element_text(size = 9,  color = "#333333"),
      axis.text        = element_text(size = 8,  color = "#555555"),
      strip.text       = element_text(face = "bold", color = "white"),
      strip.background = element_rect(fill = accent_color),
      legend.position  = "bottom",
      legend.text      = element_text(size = 8),
      legend.title     = element_text(size = 8)
    )
}

theme_set(theme_canyon())
```

# Data Collection & Preparation

```{r load data}
# Read raw lines to identify header rows
raw_lines <- readLines("https://www.waterdatafortexas.org/reservoirs/individual/canyon.csv")
```

```{r wrangle data}
canyon_raw <- read.csv(
  "https://www.waterdatafortexas.org/reservoirs/individual/canyon.csv",
  skip         = 54,
  header       = TRUE,
  comment.char = ""
)

# Aggregate to monthly
canyon_monthly <- canyon_raw %>%
  mutate(date  = as.Date(date),
         month = yearmonth(date)) %>%
  group_by(month) %>%
  summarise(
    percent_full         = mean(percent_full,         na.rm = TRUE),
    conservation_storage = mean(conservation_storage, na.rm = TRUE)
  ) %>%
  as_tsibble(index = month)

# Filter to study window: Jan 2010 – Dec 2024
canyon_monthly <- canyon_monthly %>%
  filter(month >= yearmonth("2010 Jan") &
         month <= yearmonth("2024 Dec"))

# Validation checks
nrow(canyon_monthly)       # should be 180
range(canyon_monthly$month)
canyon_monthly %>% scan_gaps()
```

## Full Series Time Plot

```{r full series plot}
canyon_monthly %>%
  autoplot(percent_full, linewidth = 1.1) +
  geom_vline(xintercept = yearmonth("2022 Jan"),
             linetype = "dashed", color = "#7B6FA0", linewidth = 0.8) +
  annotate("text", x = yearmonth("2018 Jun"), y = 85,
           label = "Training", color = "#3D7A5C", size = 3) +
  annotate("text", x = yearmonth("2023 Jun"), y = 85,
           label = "Test", color = "#A23B3B", size = 3) +
  labs(
    title    = "Canyon Lake: Percent Full (2010–2024)",
    subtitle = "Dashed line indicates train/test split | Training: Jan 2010–Dec 
    2021 | Test: Jan 2022–Dec 2024",
    y = "Percent Full (%)",
    x = NULL
  )
```

# Train / Test Split

```{r train-test split}
train_data <- canyon_monthly %>%
  filter(month <= yearmonth("2021 Dec"))   # 144 months (80%)

test_data <- canyon_monthly %>%
  filter(month >  yearmonth("2021 Dec"))   # 36 months  (20%)

cat("Training observations:", nrow(train_data), "\n")
cat("Test observations:    ", nrow(test_data),  "\n")
```

# Exploratory Analysis (Training Data Only)

### ACF / PACF

```{r acf-pacf}
train_data %>%
  gg_tsdisplay(percent_full, plot_type = "partial") +
  labs(title = "Canyon Lake: Percent Full (2010–2021)"
       )
```

### STL Decomposition

```{r stl-decomp}
train_data %>%
  model(STL(percent_full ~ trend(window = 13) +
              season(window = "periodic"))) %>%
  components() %>%
  autoplot() +
  labs(
    title    = "STL Decomposition — Canyon Lake Percent Full",
    subtitle = "Training data: Jan 2010 – Dec 2021",
    caption  = "Source: Texas Water Data (waterdatafortexas.org)"
  )
```

### Transformation & Stationarity Checks

```{r stationarity}
# Box-Cox lambda (assessed but not applied due to bounded variable)
lambda <- train_data %>%
  features(percent_full, features = guerrero) %>%
  pull(lambda_guerrero)
cat("Guerrero lambda:", round(lambda, 3), "\n")

# Seasonal differencing
train_data %>% features(percent_full, unitroot_nsdiffs)

# Non-seasonal differencing after seasonal diff
train_data %>% features(difference(percent_full, 12), unitroot_ndiffs)
```

# Model Fitting

```{r fit-models}
models <- train_data %>%
  model(
    ETS_Auto   = ETS(percent_full),
    ARIMA_Auto = ARIMA(percent_full),
    TSLM       = TSLM(percent_full ~ trend() + season()),
    Ensemble   = (ETS(percent_full) + ARIMA(percent_full) +
                    TSLM(percent_full ~ trend() + season())) / 3
  )
```

### Model Summaries

```{r model-reports}
report(models$ETS_Auto[[1]])
report(models$ARIMA_Auto[[1]])
report(models$TSLM[[1]])
```

# Residual Diagnostics

```{r arima-residuals, fig.height=4.5, fig.width=9, out.width='100%'}
# ARIMA residuals — styled (poster chart)
resid_plot <- models %>%
  select(ARIMA_Auto) %>%
  gg_tsresiduals(lag = 24) %>%
  invisible()

p1 <- resid_plot[[1]] + theme_canyon(accent_color = "#A23B3B") + labs(title = NULL)
p2 <- resid_plot[[2]] + theme_canyon(accent_color = "#A23B3B") + labs(title = NULL)
p3 <- resid_plot[[3]] + theme_canyon(accent_color = "#A23B3B") + labs(title = NULL)

(p1 / (p2 | p3)) +
  plot_layout(heights = c(1.5, 1)) +
  plot_annotation(
    title    = "Residual Diagnostics \u2014 ARIMA (Automatic)",
    subtitle = "Innovations over time, ACF, and histogram",
    theme = theme(
      plot.title      = element_text(size = 13, face = "bold", color = "#A23B3B"),
      plot.subtitle   = element_text(size = 9,  color = "#666666"),
      plot.background = element_rect(fill = "white", color = "#A23B3B", linewidth = 2)
    )
  )
```
```{r ets-residuals, fig.height=3.5, fig.width=9, out.width='100%'}
models %>%
  select(ETS_Auto) %>%
  gg_tsresiduals(lag = 24) +
  labs(title = "Residual Diagnostics — ETS (Automatic)")
```
```{r tslm-residuals, fig.height=3.5, fig.width=9, out.width='100%'}
models %>%
  select(TSLM) %>%
  gg_tsresiduals(lag = 24) +
  labs(title = "Residual Diagnostics — TSLM (Linear Regression)")
```

### Ljung-Box Test

```{r ljung-box}
models_list <- c("ETS_Auto", "ARIMA_Auto", "TSLM")

ljung_results <- augment(models) %>%
  filter(.model %in% models_list) %>%
  group_by(.model) %>%
  features(.innov, ljung_box, lag = 24, dof = 0)

ljung_results %>%
  rename(Model = .model,
         `LB Statistic` = lb_stat,
         `p-value` = lb_pvalue) %>%
  mutate(
    `LB Statistic` = round(`LB Statistic`, digits = 2),
    `p-value` = round(`p-value`, digits = 3),
    `Residuals Clean?` = ifelse(`p-value` > 0.05, "Yes", "No"))
```

> **Interpretation:** A p-value above 0.05 indicates residuals are consistent with white noise.
> ARIMA passes (p = 0.688); ETS and TSLM show significant residual autocorrelation.

# Forecasting

```{r forecast}
fc <- forecast(models, h = 36)
```

## Forecast Plot

```{r forecast-plot}
fc %>%
  autoplot(canyon_monthly, level = 95, alpha = 0.8) +
  scale_color_manual(values = model_colors) +
  scale_fill_manual(values  = model_colors) +
  labs(
    title    = "Canyon Lake Storage Forecasts: Four-Model Comparison",
    subtitle = "Forecast period: Jan 2022 – Dec 2024 | 95% prediction intervals",
    caption  = "Source: Texas Water Data (waterdatafortexas.org)",
    y = "Percent Full (%)", x = NULL, color = "Model", fill = "Model"
  )
```

# Forecast Accuracy

```{r accuracy}
accuracy_table <- fc %>%
  accuracy(test_data) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

accuracy_table %>%
  rename(Model = .model) %>%
  mutate(across(c(RMSE, MAE), ~ round(.x, 2)),
         MAPE = paste0(round(MAPE, 1), "%"))
```

# Model Selection & Justification

Although TSLM produced the lowest RMSE (28.5), its Ljung-Box p-value of approximately 0
indicates very autocorrelated residuals. This invalidates its prediction intervals, making it
unreliable for contractual planning purposes where uncertainty bounds matter as much as point
forecasts.

**The ARIMA model is selected as the best model** based on the combination of:

- Competitive forecast accuracy (RMSE = 30.4, MAE = 26.4, MAPE = 42.0%)
- Clean residuals confirmed by Ljung-Box test (p = 0.688)
- Statistically valid 95% prediction intervals suitable for drought-stage planning

For a water utility negotiating multi-year supply agreements under GBRA's tiered drought
contingency framework, the lower bound of the forecast range is operationally critical.
ARIMA is the only model that produces trustworthy uncertainty estimates.

# References

City of Kyle. (2022). *About Kyle water.* City of Kyle Water Utilities.
https://www.cityofkyle.gov/city-services/water-utilities/about-kyle-water/

Guadalupe-Blanco River Authority. (2024). *Drought contingency plan for wholesale customers.*
https://www.gbra.org/wp-content/uploads/2024/05/2024-Wholesale-Drought-Contingency-Plan_FINAL-w-Resolution.pdf

Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and practice* (3rd ed.).
OTexts. https://otexts.com/fpp3/

Rohli, E., Woolsey, N., & Sathiaraj, D. (2023). Near-term forecasting of water reservoir storage
capacities using long short-term memory. *Environmental Data Science, 2*, e30.
https://doi.org/10.1017/eds.2023.25

Stark, S. (2025, May 14). City of Kyle diversifying water sources amid pervasive drought
conditions. *KXAN Austin.*
https://www.kxan.com/news/local/hays/kyle/city-of-kyle-diversifying-water-sources-amid-pervasive-drought-conditions/

Water Data for Texas. (2024). *Canyon Lake reservoir storage data.*
https://www.waterdatafortexas.org/reservoirs/individual/canyon

